Increasing the mortality gap? Social inequality in mortality among elderly and adults in Northern Sweden 1851--2013\footnote{Session "Health, morbidity, and mortality" at the Nordic Demographic Symposium 14--16 June 2017, Turku, Finland}

### 2017-10-24 cape.Rnw
Corrections

\begin{itemize}
\item Development over \emp{time}.
  \begin{itemize}
\item Explicit or implicit assumption of an \emp{eternal} social gradient in history in many theories (Omran, Link and Phelan, Marmot, McKeown), some implying even larger in history.
\item Antonovsky – divergence followed by convergence
\end{itemize}
\item Development over the \emp{life course}
  \begin{itemize}
\item Convergence, divergence or maintenance.
\item Antonovsky – large differences in adulthood, convergence in old age
\end{itemize}
\end{itemize}

@@

Remote area in northern Sweden, dominated by agriculture until well into the 20th century.

@@

\item[Sex]
\item[Marital status]
\item[Birth date] By quarter (15 Feb, May, Aug, Nov)
\item[Death date] By day (\emp{Socialstyrelsen}, \emp{Poplink})
\item[*Death cause] from \emp{Socialstyrelsen} and \emp{Poplink}. Adapted to ICD10
\item[Parish] Urban/Rural
\item[Social class] Elite, middle, worker. Household definition 1851--1950. Individual definition 1976--2013
%\item[*Income] 1986--2005, grouped by quartiles.
%\item[*Education] 1990--2013, three categories.
\item[Period]  1851--1875, \ldots,  2008--2013 (13 periods). 
\item[Age group] Separate analyses for working active (40-65) vs retired (65-89) population
%\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).

@@

\begin{itemize}
  \item \emp{\tt sex}
  \item \emp{\tt period}: 1851--75, 1876--1900, 1901--10, etc.
  \item \emp{\tt class}
  \item \emp{\tt age group}: 40--64, 65--89.
  \item \emp{\tt cause of death}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \emp{Improved} survival characterised all social classes.
\item \emp{Late} appearance of the “modern” pattern of social differences in mortality among men
\item Distinct \emp{gender} differences in the development.
\item \emp{Large} differences between those in working age compared to the retired population, indications of \emp{status convergence}.
% \item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
\item In the long run, no clear connection in the development of income inequality and social differences in mortality.
\item This suggests an important role of gender and class related expectations on health behaviour
\end{itemize}

@@

\begin{itemize}
\item The \emp{development} of social health inequalities and the role of \emp{health-related behavior} in different phases of the transition (Antonovsky and Omran).
\item Inequality as a \emp{contextual} health risk (Wilkinson/Pickett).
\item \emp{Status syndrome} (Marmot)
\item The role of \emp{living standard} for mortality decline (McKeown).
\item \emp{Gender} and mortality (Watkins, Mckinnon, Janssens).
\item Social conditions as \emp{fundamental causes of death} (Link/Phelan).

>>>>

\begin{itemize}
\item Development over \emp{time}.
  \begin{itemize}
\item Explicit or implicit assumption of an \emp{eternal} social gradient (Omran, Link and Phelan, Marmot, McKeown), some implying even larger differences in history.
\item Antonovsky – divergence followed by convergence
\end{itemize}
\item Development over the \emp{life course}
  \begin{itemize}
\item Convergence, divergence or maintenance.
\item Antonovsky – \emp{Convergence} from adulthood to old age
\end{itemize}
\end{itemize}

@@

Remote areas in northern Sweden, dominated by agriculture until well into the 20th century.

@@

\item[Sex] Male/female
\item[Marital status] Unmarried/married/previously married
\item[Birth date] By quarter (15 Feb, May, Aug, Nov)
\item[Death date] By day (\emp{Socialstyrelsen}, \emp{Poplink})
\item[*Death cause] from \emp{Socialstyrelsen} and \emp{Poplink}. Adapted to ICD10
\item[Parish] Urban/Rural
\item[Social class] Elite, middle, worker. Household definition 1801--1950. Individual definition 1976--2013
%\item[*Income] 1986--2005, grouped by quartiles.
%\item[*Education] 1990--2013, three categories.
\item[Period]  1801--1825, \ldots,  2008--2013 (13 periods). 
\item[Age group] Separate analyses for working active (40-65) vs retired (65-89) population
%\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).

@@

\begin{itemize}
  \item \emp{\tt sex}
  \item \emp{\tt period}: 1801--25, 1826--1850, ... 1976--1980, etc.
  \item \emp{\tt class}
  \item \emp{\tt age group}: 40--64, 65--89.
  \item \emp{\tt cause of death}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \emp{Improved} survival characterised all social classes.
\item \emp{Late} appearance of the “modern” pattern of social differences in mortality among men
\item Distinct \emp{gender} differences in the development.
\item \emp{Large} differences between those in working age and the retired population, indications of \emp{status convergence}.
% \item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
\item No clear connection in the development of income inequality and social differences in mortality.
\item The results suggest an important role of gender and class related expectations on health behaviour
\end{itemize}

@@

\begin{itemize}
\item The role of \emp{health-related behavior} in different phases of the transition (Antonovsky and Omran).
\item Inequality as a \emp{contextual} health risk (Wilkinson/Pickett).
\item \emp{Status syndrome} (Marmot)
\item The role of \emp{living standard} for mortality decline (McKeown).
\item \emp{Gender} and mortality (Watkins, Mckinnon, Janssens).
\item Social conditions as \emp{fundamental causes of death} (Link/Phelan).
